---
title: "AMS597_Group-2_Regression"
author: "Ishaan Reddy"
date: "2026-03-29"
output: html_document
---

```{r}
# Load data
data_spotify <- read.csv("/Users/ishaa/Documents/SBU/SBU_Sem-2/Stats Computing/Project/dataset.csv")

# Preprocessing
data_spotify$duration_s <- data_spotify$duration_ms / 1000
data_spotify$duration_ms <- NULL
data_spotify$explicit <- as.logical(data_spotify$explicit)
data_spotify$X <- NULL

data_spotify_merged <- aggregate(track_genre ~ track_id + track_name + artists + 
                                   album_name + popularity + duration_s + explicit + 
                                   danceability + energy + key + loudness + mode + 
                                   speechiness + acousticness + instrumentalness + 
                                   liveness + valence + tempo + time_signature, 
                                 data = data_spotify, 
                                 FUN = function(x) paste(sort(unique(x)), collapse = ", "))

data_spotify_clean <- data_spotify_merged[
  data_spotify_merged$duration_s > 0 &
  data_spotify_merged$duration_s <= 600 &
  data_spotify_merged$loudness >= -35 &
  data_spotify_merged$tempo != 0, 
]
```


```{r}
head(data_spotify_clean)

#Plan:
#Split data into train/test
#Run Ridge regression → shows all features matter with shrinkage
#Run Lasso regression → automatically drops unimportant features
#Compare both using RMSE and R²
#Plot coefficient paths and variable importance
```


```{r}
# Load required libraries
install.packages("glmnet")
library(glmnet)
```


```{r}
# Select predictors and response
X <- model.matrix(popularity ~ danceability + energy + loudness + speechiness +
                    acousticness + instrumentalness + liveness + valence +
                    tempo + duration_s + explicit, 
                  data = data_spotify_clean)[, -1]  # Remove intercept column

y <- data_spotify_clean$popularity

set.seed(42)
train_idx <- sample(1:nrow(X), 0.8 * nrow(X))
X_train <- X[train_idx, ]
X_test  <- X[-train_idx, ]
y_train <- y[train_idx]
y_test  <- y[-train_idx]
```



```{r}
# Ridge Regression
set.seed(42)
ridge_cv <- cv.glmnet(X_train, y_train, alpha = 0)
best_lambda_ridge <- ridge_cv$lambda.min
plot(ridge_cv)

# Lasso Regression
set.seed(42)
lasso_cv <- cv.glmnet(X_train, y_train, alpha = 1)
best_lambda_lasso <- lasso_cv$lambda.min
plot(lasso_cv)
```




```{r}
# Ridge predictions
ridge_pred <- predict(ridge_cv, s = best_lambda_ridge, newx = X_test)
ridge_rmse <- sqrt(mean((ridge_pred - y_test)^2))
ridge_r2   <- 1 - sum((ridge_pred - y_test)^2) / sum((y_test - mean(y_test))^2)

# Lasso predictions
lasso_pred <- predict(lasso_cv, s = best_lambda_lasso, newx = X_test)
lasso_rmse <- sqrt(mean((lasso_pred - y_test)^2))
lasso_r2   <- 1 - sum((lasso_pred - y_test)^2) / sum((y_test - mean(y_test))^2)

# Print results
cat("Ridge - RMSE:", round(ridge_rmse, 4), " R²:", round(ridge_r2, 4), "\n")
cat("Lasso - RMSE:", round(lasso_rmse, 4), " R²:", round(lasso_r2, 4), "\n")
```


```{r}
# Ridge coefficients
ridge_coef <- coef(ridge_cv, s = best_lambda_ridge)
print("Ridge Coefficients:")
print(ridge_coef)

# Lasso coefficients
lasso_coef <- coef(lasso_cv, s = best_lambda_lasso)
print("Lasso Coefficients:")
print(lasso_coef)
```



```{r}
install.packages("xgboost")
library(xgboost)
```


```{r}
# XGBoost requires matrix format
dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dtest  <- xgb.DMatrix(data = X_test,  label = y_test)

set.seed(42)
xgb_model <- xgboost(
  x = X_train,
  y = y_train,
  nrounds = 200,
  max_depth = 6,
  learning_rate = 0.1,
  subsample = 0.8,
  colsample_bytree = 0.8,
  objective = "reg:squarederror"
)
```

```{r}
xgb_pred <- predict(xgb_model, newdata = X_test)
xgb_rmse <- sqrt(mean((xgb_pred - y_test)^2))
xgb_r2   <- 1 - sum((xgb_pred - y_test)^2) / sum((y_test - mean(y_test))^2)

cat("XGBoost - RMSE:", round(xgb_rmse, 4), " R²:", round(xgb_r2, 4), "\n")
```


```{r}
importance_matrix <- xgb.importance(model = xgb_model)
xgb.plot.importance(importance_matrix, top_n = 11,
                    main = "XGBoost Feature Importance")
```
**Interpretaion of findings and conclusion to the research question "Which audio features best predict a track's popularity?"**

Research Question 1 asked which audio features best predict a track's popularity. We first applied Lasso and Ridge regression as baseline models, both yielding an R² of 0.03 and RMSE of 20.38, indicating that linear relationships between audio features and popularity are negligible. This motivated the use of XGBoost, a gradient boosting model capable of capturing non-linear patterns, which improved performance to an R² of 0.17 and RMSE of 18.86 — a five-fold improvement over the linear baselines.

The XGBoost feature importance plot revealed that acousticness, duration, danceability, and instrumentalness were the strongest predictors of popularity, while explicit content contributed the least. Despite the improvement, the overall R² of 0.17 suggests that audio features alone have limited predictive power over popularity, which is expected — popularity is also heavily influenced by external factors such as artist recognition, marketing, and cultural trends that are not captured in audio-based features alone.




