---
title: "AMS597 Group 2 - Clustering Analysis"
output: pdf_document
date: "2026-04-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Preprocessing

```{r}
data_spotify <- read.csv("dataset.csv")
head(data_spotify)
```




```{r}
print("Structure and Data Types:")
str(data_spotify)
```


```{r}
#Check if any columns have null values
colSums(is.na(data_spotify))
```


```{r}
#Convert the duration from ms to s
data_spotify$duration_s <- data_spotify$duration_ms / 1000

# Drop the original duration_ms column
data_spotify$duration_ms <- NULL

# Convert Explicit column to Boolean values from chr
data_spotify$explicit <- as.logical(data_spotify$explicit)

#Drop index column
data_spotify$X <- NULL

head(data_spotify)
```

```{r}
sum(duplicated(data_spotify$track_id))
```

```{r}
# Merge rows by track_id, combining genres into a single comma-separated string
data_spotify_merged <- aggregate(track_genre ~ track_id + track_name + artists + 
                                   album_name + popularity + duration_s + explicit + 
                                   danceability + energy + key + loudness + mode + 
                                   speechiness + acousticness + instrumentalness + 
                                   liveness + valence + tempo + time_signature, 
                                 data = data_spotify, 
                                 FUN = function(x) paste(sort(unique(x)), collapse = ", "))

# Verify
nrow(data_spotify_merged) 
```

```{r}
print("Summary Statistics:")
summary(data_spotify_merged)
```

```{r}
# Boxplots to visualize outliers
par(mfrow = c(2, 3))
boxplot(data_spotify_merged$popularity,   main = "Popularity")
boxplot(data_spotify_merged$duration_s,  main = "Duration")
boxplot(data_spotify_merged$loudness,     main = "Loudness")
boxplot(data_spotify_merged$tempo,        main = "Tempo")
boxplot(data_spotify_merged$speechiness,  main = "Speechiness")
boxplot(data_spotify_merged$liveness,     main = "Liveness")
```

```{r}
# Check extreme durations
range(data_spotify_merged$duration_s)

# How many songs are over 10 minutes (600 seconds) and equal to 0 seconds?
sum(data_spotify_merged$duration_s > 600)
sum(data_spotify_merged$duration_s == 0)
```

```{r}
# Check extreme loudness
range(data_spotify_merged$loudness)

# How many songs below -35 dB?
sum(data_spotify_merged$loudness < -35)
```
```{r}
# How many songs with tempo of 0?
sum(data_spotify_merged$tempo == 0)
```

```{r}
#Data cleaning based on the above explanation
data_spotify_clean <- data_spotify_merged[
  data_spotify_merged$duration_s > 0 &
  data_spotify_merged$duration_s <= 600 &
  data_spotify_merged$loudness >= -35 &
  data_spotify_merged$tempo != 0, 
]

# Verify
nrow(data_spotify_merged)  # Count before
nrow(data_spotify_clean)   # Count after
```


# Clustering Research Question

The goal of this analysis is to determine whether natural groupings of Spotify tracks emerge from their audio features and whether these clusters align with genre labels.

```{r}
# Clustering Analysis

# audio features for clustering
cluster_vars <- data_spotify_clean[, c(
  "danceability",
  "energy",
  "loudness",
  "speechiness",
  "acousticness",
  "instrumentalness",
  "liveness",
  "valence",
  "tempo"
)]

str(cluster_vars)
```
```{r}
# Standardize audio features before clustering
cluster_scaled <- scale(cluster_vars)

# Check result
summary(cluster_scaled)
```






```{r}
# Elbow Method to find optimal K
set.seed(597)
wss <- numeric(10)

for (k in 1:10) {
  kmeans_model <- kmeans(
    cluster_scaled,
    centers = k,
    nstart = 20,
    iter.max = 50
  )
  wss[k] <- kmeans_model$tot.withinss
}

plot(1:10, wss, type = "b",
     xlab = "Number of Clusters (K)",
     ylab = "Within-Cluster Sum of Squares",
     main = "Elbow Method")
```
```{r}
# Final K-means clustering
set.seed(597)

kmeans_final <- kmeans(
  cluster_scaled,
  centers = 4,
  nstart = 25,
  iter.max = 50
)

# Add cluster labels to dataset
data_spotify_clean$cluster <- kmeans_final$cluster

# Check cluster sizes
table(data_spotify_clean$cluster)
```


```{r}
# Use PCA for visualization (2D plot)
pca_vis <- prcomp(cluster_scaled)

# Create a dataframe for plotting
pca_df <- data.frame(
  PC1 = pca_vis$x[,1],
  PC2 = pca_vis$x[,2],
  cluster = as.factor(data_spotify_clean$cluster)
)

# Plot clusters
plot(pca_df$PC1, pca_df$PC2,
     col = pca_df$cluster,
     pch = 16,
     main = "K-means Clusters (K = 4)",
     xlab = "PC1",
     ylab = "PC2")
```

```{r}
# Mean values of each feature per cluster
cluster_summary <- aggregate(cluster_vars,
                             by = list(cluster = data_spotify_clean$cluster),
                             FUN = mean)

cluster_summary
```

```{r}
# Compare clusters with genres
# table(data_spotify_clean$cluster, data_spotify_clean$track_genre)
```

```{r}
# ============================
# Cluster vs Genre (Top Genres)
# ============================

library(dplyr)

top_genres <- data_spotify_clean %>%
  group_by(cluster, track_genre) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(cluster, desc(count)) %>%
  group_by(cluster) %>%
  slice_head(n = 5)
top_genres
```

# Conclusion

The clustering analysis identified four distinct groups of Spotify tracks based on their audio features. Cluster 2 represents softer and more acoustic tracks, while Clusters 3 and 4 capture more energetic and less acoustic songs. Cluster 1 shows relatively higher speechiness and vocal-heavy characteristics. Overall, the clusters are musically meaningful and partially align with genre labels, although genre is not explained perfectly by audio features alone.


















